\chapter{Introduction}\label{ch:intro}

Backpropagation is the engine of modern deep learning, yet it is widely regarded as
biologically implausible. Training a network by the chain rule requires a global backward pass
that transports the transpose of each weight matrix back through the network (the
\emph{weight-transport problem}), that locks every layer's update until the full forward and
backward sweeps complete, and that carries an error signal which is mechanistically distinct
from the activity used in the forward pass. Cortical circuits exhibit none of these properties:
they learn with local signals, in place, and apparently without a separate backward channel.
This mismatch has motivated a sustained search for \emph{local} learning rules that could, in
principle, be implemented by biological or neuromorphic substrates while still solving the
credit-assignment problem that backpropagation solves globally.

Predictive Coding (PC) is among the most developed of these alternatives. Originating as a
theory of cortical function \citep{rao1999predictive} and later cast in the language of the
free-energy principle \citep{bogacz2017tutorial}, PC frames perception and learning as the
minimisation of a layer-wise, precision-weighted prediction-error energy. Inference is an
iterative \emph{settling} process that relaxes the network's internal states toward a
configuration of low prediction error, after which a purely local Hebbian rule updates the
weights. PC is attractive precisely because it is local, and it has accumulated a set of
appealing claims: that it approximates backpropagation \citep{whittington2017approximation};
that it can learn from fewer examples; that it forgets less in continual-learning settings; and
that the very same machinery doubles as a generative model. If even some of these held robustly,
PC would be more than a biological curiosity.

The difficulty is that the evidence is inconsistent. Reported advantages appear in some studies
and regimes and vanish in others, and comparisons across the literature are rarely controlled
for the many degrees of freedom that separate a PC network from its backpropagation baseline
\citep{zahid2023curvature}. At the same time, PC's iterative inference makes it markedly more
expensive to run than a feedforward pass, and the engineering question of \emph{where that cost
actually goes} on modern hardware has received little attention. This thesis takes a deliberately
sober stance toward both issues. It does not attempt a breakthrough; it builds a clean,
from-scratch PC implementation and uses it to ask careful, falsifiable questions, reporting the
answers---several of them null---honestly.

\section{Motivation}

The motivation for this work is threefold. \emph{Scientifically}, the inconsistency in the
PC-versus-backpropagation literature is itself a problem worth addressing: if the reported
advantages are real, a controlled study should recover them, and if they are artefacts, a
controlled study should expose the artefacts. Either outcome is informative. \emph{From a
systems perspective}, PC's settling loop is an unusual workload---many small, sequential,
launch-bound operations rather than a few large matrix multiplications---and characterising its
cost is both practically useful and a natural bridge to the wider literature on accelerating
iterative, fixed-point computations. \emph{Pedagogically}, implementing PC from first
principles, with no reliance on automatic differentiation for the learning rule, forces every
equation to be understood and verified, which is the surest route to separating what PC
genuinely does from what a high-level framework does on its behalf.

A recurring theme throughout the thesis is that fair comparison is hard, and that most apparent
PC advantages dissolve once the comparison is made fair. Rather than treat this as a
disappointment, we treat methodological honesty---naming the confounds, controlling them,
reporting confidence intervals and seed counts, and declining to call a marginal effect
significant---as the central contribution.

\section{Problem Statement and Research Questions}

We organise the work around three research questions, each of which maps to one of the
contributions below.

\begin{description}
  \item[RQ1 (Science).] Under a comparison that matches architecture, initialisation, data, and
    loss function, tunes each method's learning rate fairly, and reports uncertainty over
    multiple seeds, does Predictive Coding actually outperform backpropagation at MNIST scale---on
    accuracy, robustness, sample efficiency, or continual learning---and if not, what explains
    the advantages reported elsewhere?
  \item[RQ2 (Systems).] What dominates the cost of PC inference on a GPU, and can that cost be
    reduced by a dedicated implementation of the settling loop?
  \item[RQ3 (Capability).] Can a single from-scratch PC network act as one mechanism for many
    tasks---classification, generation, inpainting, and anomaly detection---and under what
    conditions does this property actually emerge?
\end{description}

These questions are intentionally answerable at the scale we can afford to study rigorously. They
trade breadth for the ability to give each answer with controlled experiments and honest
uncertainty, rather than gesturing at trends.

\section{Contributions}

The thesis makes three contributions, aligned with the research questions.

\begin{description}
  \item[(A) A fused CUDA settling kernel and a characterisation of PC's inference cost.] We
    contribute, to our knowledge, the first hand-written fused CUDA kernel for the PC settling
    loop. The persistent, states-resident, single-launch \emph{technique} is established prior art
    in other domains; our contribution is its transfer to PC together with the measurement that
    PC inference at MNIST scale is launch-overhead-bound. The kernel yields a $3.2\times$ speedup
    at small batch, generalises to networks of arbitrary depth, and is fast enough to serve as
    the actual backend of one of our own studies (\Cref{ch:method,ch:experiments}).
  \item[(B) A confound-controlled fair-comparison protocol and a parity finding.] We design a
    PC-versus-backpropagation protocol that holds everything but the learning rule fixed, and find
    parity across every axis we test---including an architecture-faithful replication of an exact
    protocol from the literature. We isolate three confounds that manufacture spurious
    ``PC advantages'' in naive comparisons, and we describe an adversarial verification procedure
    that caught a literature-framing error in our own draft before it reached this document
    (\Cref{ch:experiments,ch:discussion}).
  \item[(C) A working generative PC, plus two implemented extensions.] We show that the
    ``one mechanism, many tasks'' property is real only for a \emph{generatively} oriented PC, and
    demonstrate classification, generation, inpainting, and anomaly detection from a single
    settling rule. We additionally implement incremental PC and an Elastic Weight Consolidation
    baseline (\Cref{ch:method,ch:experiments}).
\end{description}

All code, the test suite, the figures, and the raw experiment artifacts are released publicly so
that every number in this thesis can be reproduced.

\section{Scope and Non-Goals}

This is a study at MNIST and FashionMNIST scale, with shallow multilayer perceptrons (two to four
hidden layers) and the vanilla State-Optimization formulation of PC. It is explicitly \emph{not}
a claim about Transformer-scale models, nor a state-of-the-art result, nor a refutation of any
prior work in that work's own regime. The honest framing of the project is that a from-scratch
MNIST PC network is primarily an exercise in \emph{understanding and clean reproduction}, and
that whatever novelty it carries comes from a small number of focused, defensible
angles---a systems measurement, a methodological checklist, and a capability distinction---rather
than from a paradigm shift. \Cref{ch:limitations} states the boundaries of every claim in full.

\section{Outline of the Thesis}

\Cref{ch:background} develops the State-Optimization formulation of PC: the prediction-error
energy, the settling dynamics and the local weight update, the precise sense in which PC relates
to backpropagation, and the orientation distinction between discriminative and generative PC.
\Cref{ch:related} surveys the prior art---PC libraries, Equilibrium Propagation as the closest
algorithmic relative, systems work on fixed-point inference, and the continual-learning
literature---and positions the thesis's claims, in particular grounding the systems-novelty claim
in a direct source audit. \Cref{ch:method} describes what was built: the from-scratch network and
its single training interface, the gradient checks against autograd, the fused CUDA kernel, the
fair-comparison protocol, the continual-learning setups including the exact replication, the
generative model, and the incremental-PC and EWC extensions. \Cref{ch:experiments} reports the
measurements---the kernel benchmark and launch-count analysis, the fair PC-versus-BP study and the
three confounds, the budget-swept exact replication, the generative results, and the extensions.
\Cref{ch:discussion} interprets the findings, develops the confound checklist and the systems
insight as transferable contributions, and reflects on the adversarial-verification methodology.
\Cref{ch:limitations} states the limitations and the concrete future work that would extend the
claims, and \Cref{ch:conclusion} concludes.
