\chapter{Background}\label{ch:background}

This chapter develops the theoretical machinery that the rest of the thesis rests upon.
It is deliberately self-contained and pedagogical: we begin from the familiar feedforward
multilayer perceptron (MLP) trained by backpropagation, introduce the predictive-coding (PC)
energy that reinterprets a network as an inference problem, derive the State-Optimization (SO)
settling dynamics and the local Hebbian weight rule in full, and only then turn to the precise
relationship between PC and backpropagation. We restrict ourselves throughout to \emph{theory}.
Our concrete implementation choices, the libraries that preceded us, and our empirical results
are deferred to \Cref{ch:method,ch:related,ch:experiments} respectively. The aim here is that a
reader who has never seen predictive coding should, by the end of the chapter, be able to read
the settling and update equations and understand exactly what each term does and why.

\section{Notation and the MLP Baseline}

We fix notation now and use it consistently for the remainder of the thesis. Consider a network
of $n+1$ layers indexed $0,1,\dots,n$. Layer $0$ holds the input and layer $n$ holds the output;
the layers in between are hidden. We write the \emph{state} (the vector of unit activities) of
layer $k$ as $s_k$, the weight matrix mapping layer $i$ to layer $i+1$ as $W_i$, and the
corresponding bias as $b_i$. The element-wise activation function is $\phi$ (we use $\tanh$ as the
default throughout, with the sigmoid reserved for one faithfulness replication), and $\phi'$
denotes its derivative. Layer sizes for our running MNIST example are
$784 \to 256 \to 256 \to 10$: a $784$-dimensional pixel vector, two hidden layers of width $256$,
and a $10$-dimensional output read as a one-hot class label \citep{lecun1998mnist}.

\subsection{The feedforward MLP}

The ordinary feedforward MLP computes its activities by a single forward sweep. Writing the
pre-activation of layer $i+1$ as
\begin{equation}
  \mathit{pred}_{i+1} = W_i\,\phi(s_i) + b_i ,
  \label{eq:prediction}
\end{equation}
the MLP simply \emph{assigns} $s_{i+1} = \mathit{pred}_{i+1}$ for every layer in turn, starting
from the clamped input $s_0 = x$ and ending at the output $s_n$. We deliberately call the
right-hand side of \cref{eq:prediction} a \emph{prediction}: it is the higher layer's best guess,
given $s_i$, of what the activity of layer $i+1$ ought to be. In a plain MLP that guess is taken
as gospel and copied into $s_{i+1}$. The central conceptual move of predictive coding, developed
below, is to stop treating the prediction as definitive and instead let the actual activity
$s_{i+1}$ disagree with it; the discrepancy is precisely what drives learning.

\subsection{Backpropagation as the reference}

Given a differentiable scalar loss $\mathcal{L}(s_n, y)$ comparing the output to a target $y$,
backpropagation (BP) computes the gradient of $\mathcal{L}$ with respect to every weight by the
chain rule, propagating an error signal from the output back toward the input. Two properties of
this procedure matter for what follows. First, the weight update for layer $i$ depends on a
quantity transported from \emph{all} downstream layers: the backward pass threads the transposed
weight matrices $W_{i+1}^\top, W_{i+2}^\top, \dots$ in sequence, so the signal reaching layer $i$
has been routed through the entire upper part of the network. This non-locality is the
``weight-transport problem'' often cited as a reason BP is biologically implausible. Second, BP
changes the weights \emph{first}; any change in the network's activities is a downstream
consequence of the next forward pass, not part of the learning step itself. We will see that
predictive coding inverts both of these properties: its weight update is strictly local, and it
changes activities \emph{before} it touches the weights. Throughout the thesis, a matched-size MLP
trained by BP under PyTorch autograd \citep{paszke2019pytorch} serves as the reference against
which PC is measured; the discipline required to make that comparison fair is the subject of
\Cref{sec:protocol}.

\section{Predictive Coding: The Energy Formulation}

Predictive coding originates in computational neuroscience as a theory of cortical processing
\citep{rao1999predictive}: higher cortical areas send \emph{predictions} of activity down to lower
areas, and only the residual \emph{prediction error}---the part of the lower activity that the
prediction failed to explain---is sent back up. Learning and perception both reduce to making
those prediction errors small. \citet{bogacz2017tutorial} recast this picture in the language of
the variational free-energy framework and gave the tutorial derivation we follow here;
\citet{millidge2021review} survey the multi-layer formulation in detail. The result is a network
that is defined not by a forward function but by a single scalar \emph{energy}.

\subsection{Prediction errors and the energy}

For each layer $k \in \{1,\dots,n\}$ we define the prediction error as the difference between the
layer's actual state and the top-down prediction of it from the layer below,
\begin{equation}
  \varepsilon_k = s_k - \mathit{pred}_k = s_k - \bigl(W_{k-1}\,\phi(s_{k-1}) + b_{k-1}\bigr).
  \label{eq:error}
\end{equation}
The network's \emph{energy} is the precision-weighted sum of squared prediction errors. In full
generality each layer carries a precision matrix $\Pi_k$ (the inverse covariance of the assumed
Gaussian noise on that layer), giving
\begin{equation}
  E = \tfrac{1}{2}\sum_{k=1}^{n} \varepsilon_k^\top \Pi_k\, \varepsilon_k .
  \label{eq:free-energy}
\end{equation}
This is exactly the Gaussian form of the variational free energy of the free-energy principle,
wired layer-by-layer into the network \citep{bogacz2017tutorial,millidge2021review}. It is worth
stressing what \cref{eq:free-energy} \emph{is}: a single number that measures how badly the
network's own top-down model fails to explain its activities. Perception and learning are then two
ways of pushing $E$ downhill---by adjusting the activities $s_k$ (inference) or by adjusting the
weights $W_i$ (learning), respectively.

\subsection{Isotropic precision as the modern default}

When every precision is the identity, $\Pi_k = I$, the quadratic form in \cref{eq:free-energy}
collapses to a plain sum of squared errors,
\begin{equation}
  E = \tfrac{1}{2}\sum_{k=1}^{n} \lVert \varepsilon_k \rVert^2 .
  \label{eq:sse-energy}
\end{equation}
This \emph{isotropic} (precision-free) energy is the formulation we adopt for the entire thesis,
and it is the mainstream choice in modern PC research on image classification: contemporary
benchmark libraries fix the covariance to the identity explicitly rather than learning it
\citep{pinchetti2024pcx,millidge2024jpc}. Non-identity precision is a genuine and theoretically
well-motivated feature---a per-layer scalar precision schedule is the lever that helps very deep
PC networks train at all \citep{qi2025settling}---but its benefits accrue mainly beyond roughly
seven layers and are orthogonal to the questions this thesis asks. For the shallow MNIST-scale
networks studied here, $\Pi_k = I$ is both adequate and standard, and we treat
\cref{eq:sse-energy} as the energy of record. We note in passing that an attentive reader will
have spotted a notational subtlety: \cref{eq:error} indexes predictions in the
\emph{feedforward} direction (layer $i$ predicts layer $i+1$), whereas the neuroscience literature
often writes them top-down. The two conventions are identical under an index flip
\citep{millidge2021review}; we use the feedforward convention consistently because it is the one
the settling dynamics and the kernel of \Cref{sec:kernel} are written in.

\section{State-Optimization PC: Settling and the Local Update}\label{sec:so-pc}

We now give the exact algorithm that the rest of the thesis implements, benchmarks, and verifies.
``State Optimization'' (SO) is the standard predictive-coding formulation: it minimizes the energy
\cref{eq:sse-energy} by gradient descent, alternating two phases. In the \emph{inference} (or
\emph{settling}) phase the weights are held fixed and the free states are relaxed toward an energy
minimum; in the \emph{learning} phase the states are held at that minimum and the weights take a
single gradient step. Because both phases are gradient descent on the \emph{same} energy, the
whole scheme is a coordinate-descent on $E$ over the joint variables $(s,W)$. The presentation
here follows the algorithm exactly as we implement it (cf.\ \texttt{pcn/settling.py} and
\texttt{pcn/model.py}); it is the SO algorithm distilled from \citet{goemaere2025epc}.

\subsection{Clamping conventions}

Before the dynamics we must say which states are pinned and which are free, because this is the
single point most easily got wrong. The input layer $s_0$ is always clamped to the image $x$.
During \emph{training} the output layer $s_n$ is additionally clamped to the one-hot label $y$, so
only the hidden states settle; the equilibrium configuration is then read off to update the
weights. At \emph{test} time only the input is clamped, the output is left free, and the predicted
class is the $\arg\max$ over the settled output state---we let the label node settle rather than
supplying it, which is precisely the difference between classifying and merely reconstructing.
A third, \emph{generative} convention clamps the label at the top and lets the input settle at the
bottom, ``painting'' an image; whether this produces digits or noise depends on the training
objective, a subtlety we return to in \cref{sec:gen-disc}.

\subsection{Feedforward initialization}

Settling needs a starting point. We initialize the states by a single feedforward sweep,
\begin{equation}
  s_0 = x, \qquad s_i = \mathit{pred}_i = W_{i-1}\,\phi(s_{i-1}) + b_{i-1}
  \quad\text{for } i = 1,\dots,n,
\end{equation}
computed sequentially layer by layer. This choice is not arbitrary: copying each prediction into
the corresponding state makes every internal prediction error \cref{eq:error} exactly zero, so the
network starts at the bottom of the ``internal'' part of the energy landscape with all the
residual energy concentrated at the output. As \citet{goemaere2025epc} observe, this feedforward
heuristic works well because it coincides with the first step of the settling dynamics itself.

\subsection{The settling dynamics}

In the inference phase the weights are fixed and we descend the energy in the states. The gradient
of \cref{eq:sse-energy} with respect to a free state $s_k$ has two contributions, and the
structure of these two terms is the entire conceptual content of PC inference. Differentiating,
\begin{equation}
  \nabla_{s_k} E \;=\; \underbrace{\varepsilon_k}_{\text{own error}}
  \;-\; \underbrace{\phi'(s_k)\odot\bigl(W_k^\top\,\varepsilon_{k+1}\bigr)}_{\text{error caused below}},
  \label{eq:state-grad}
\end{equation}
where $\odot$ is the element-wise (Hadamard) product. The state then takes a step downhill,
\begin{equation}
  s_k \;\leftarrow\; s_k - \eta_x\,\nabla_{s_k} E
  \;=\; s_k - \eta_x\Bigl(\varepsilon_k - \phi'(s_k)\odot\bigl(W_k^\top\,\varepsilon_{k+1}\bigr)\Bigr),
  \label{eq:state-update}
\end{equation}
with $\eta_x$ the state learning rate (also written $\lambda$). The two terms have a clean
interpretation. The first, $\varepsilon_k$, is the error that layer $k$ itself incurs against the
prediction coming from below; moving $s_k$ to reduce it makes the layer better match what was
expected of it. The second term is the error that $s_k$ \emph{causes} in the layer above:
$\varepsilon_{k+1}$ depends on $s_k$ through $\mathit{pred}_{k+1} = W_k\phi(s_k) + b_k$, and the
chain rule produces the transposed weight $W_k^\top$ gated by the local activation derivative
$\phi'(s_k)$. Each unit therefore moves so as to simultaneously reduce its own mismatch and the
mismatch it propagates upward. Iterating \cref{eq:state-update} for $t = 1,\dots,T$ steps lets the
network ``settle'' into an energy minimum. Two boundary conditions complete the picture: a clamped
state (input always; output during training) is simply not updated, and for the free output layer
$k=n$ there is no layer above, so the second term vanishes and $\nabla_{s_n} E = \varepsilon_n$.

A note on implementation that is also conceptually load-bearing: within a single settling step the
updates over layers are computed in \emph{parallel} (a Jacobi sweep), because every
$\varepsilon_k$ and every $\nabla_{s_k} E$ depends only on quantities from the previous step. Only
the loop over the $T$ steps and the feedforward initialization are inherently sequential. This
parallel-over-layers, sequential-over-time structure is exactly what the fused settling kernel of
\Cref{sec:kernel} exploits, and the state gradient \cref{eq:state-grad} is the quantity our tests
verify against automatic differentiation.

\subsection{The signal-decay caveat}

A subtlety of the settling dynamics deserves mention, both because it bounds the regime in which
SO is well behaved and because it motivates work discussed in \Cref{ch:related}. At feedforward
initialization all internal errors are zero and only the output error is non-zero. The settling
update \cref{eq:state-update} then carries that signal backward through the network at a rate of
roughly one layer per step, attenuated at each step by the factor $\eta_x < 1$ needed for
stability. The error norm reaching a layer $i$ steps below the output therefore scales
binomially, $\lVert\varepsilon_{n-i}\rVert \propto \eta_x^{\,i}(1-\eta_x)^{\,t-i}$, an exponential
decay with depth \citep{goemaere2025epc}. For small $\eta_x$ this drives the deep-layer errors
below the floating-point floor within a handful of steps, leaving deep layers effectively
untrained. For the shallow MNIST networks of this thesis the effect is negligible---two or three
hidden layers are simply not deep enough---so SO works well; the decay only becomes a genuine
obstacle at depths well beyond what we study. It does, however, refine the folklore on how many
settling steps are needed: the safe lower bound is $T \ge L$ for an $L$-layer network
\citep{qi2025settling}, with the upper bound governed by the decay rather than by a fixed multiple
of depth.

\subsection{The local Hebbian weight update}

Once the states have settled to (approximate) equilibrium we hold them fixed and take one gradient
step in the weights. The gradient of \cref{eq:sse-energy} with respect to $W_i$ is, by
\cref{eq:error},
\begin{equation}
  \nabla_{W_i} E = -\,\varepsilon_{i+1}\,\phi(s_i)^\top ,
\end{equation}
so the weight update is
\begin{equation}
  W_i \;\leftarrow\; W_i + \eta_w\,\varepsilon_{i+1}\,\phi(s_i)^\top ,
  \label{eq:weight-update}
\end{equation}
with $\eta_w$ the weight learning rate (kept separate from $\eta_x$ throughout). This is the heart
of the whole approach. The update for $W_i$ requires \emph{only} the prediction error of the layer
above, $\varepsilon_{i+1}$, and the activity of the layer below, $\phi(s_i)$---both quantities
local to the synapse $W_i$ connects. It is a Hebbian outer product of a presynaptic activity and a
postsynaptic error: there is no global backward pass, and no chain of transposed weight matrices
threaded across the whole network \citep{whittington2017approximation,bogacz2017tutorial}. The
weight-transport problem that makes backpropagation biologically implausible simply does not
arise. This locality is what the term ``local learning'' in the title of this thesis refers to,
and recovering it on a parallel accelerator without sacrificing speed is the engineering problem
\Cref{sec:kernel} attacks.

\section{Relation to Backpropagation}

Predictive coding and backpropagation are not unrelated algorithms that happen to solve the same
task; under a precise and well-studied assumption, the local PC update of \cref{eq:weight-update}
\emph{computes the same gradient} that backpropagation would. Understanding exactly when this holds
is essential to interpreting our comparison results honestly, so we state it carefully.

\subsection{The fixed-prediction equivalence}

The foundational result is due to \citet{whittington2017approximation}: a supervised predictive
coding network with value nodes, error nodes, transposed-weight feedback gated by the activation
derivative, and the local Hebbian rule \cref{eq:weight-update} \emph{approximates} the
backpropagation gradient, and does so without any non-local plasticity. \citet{millidge2022backprop}
generalized this from layered MLPs to arbitrary computation graphs. The mechanism is the
following. If the network is at the feedforward state---the configuration in which every internal
error is zero and only the output error is non-zero---then the settling update
\cref{eq:state-update}, run for a short time, transports the output error backward through exactly
the transposed weights $W_k^\top$ gated by $\phi'(s_k)$ that backpropagation uses. Under this
``fixed-prediction'' assumption, the prediction errors $\varepsilon_k$ at the layers come to equal
the backpropagated error signals, and the local update \cref{eq:weight-update} reproduces the BP
weight gradient. In the limiting case the correspondence is exact \citep{whittington2017approximation}.
A PC network can therefore learn the same thing as an MLP, but by minimizing free energy rather
than by applying the chain rule.

\subsection{Prospective configuration is the equilibrium, not a separate algorithm}

This equivalence is sometimes obscured by the framing of ``prospective configuration''
\citep{song2024prospective}. In that framing, a PC network first changes its neural activity
across the network so that the output neurons better predict the target, and \emph{then}
consolidates that activity change into the weights---the reverse of backpropagation's
weights-first, activity-follows ordering. It is important to be clear that prospective
configuration is not a new learning rule grafted onto PC: it is simply a name for the behavior of
standard energy-based PC when the activities are allowed to relax to equilibrium before the weights
move. The inference-before-learning two-phase structure of \cref{sec:so-pc} \emph{is} prospective
configuration. The substantive empirical claim attached to the term---that this ordering helps in
regimes such as online, continual, and few-shot learning---is a separate matter that we test
directly and report, with appropriate caveats, in \Cref{sec:exp-pcbp,sec:exp-song}.

\section{When Does PC Equal BP?}

Given that PC can recover the BP gradient under the fixed-prediction assumption, the natural
question---and the one that frames our first research question---is the converse: \emph{when does
PC do something genuinely different from BP?} The answer determines whether predictive coding is
merely a biologically flavored re-derivation of backpropagation or a distinct learning algorithm
with its own behavior, and it must be stated with care to avoid overclaiming in either direction.

\subsection{The provable equivalence regimes are narrow}

The conditions under which PC provably converges to BP are restrictive. The most recent analysis,
\citet{innocenti2026limits}, shows that the PC energy with equilibrated activities converges to
the quadratic backpropagation loss and yields the same gradients---but only for \emph{linear}
residual networks, in the regime of \emph{width far greater than depth}, and only when the
activity equilibrium is genuinely reached. This is emphatically \emph{not} an unconditional
``PC $=$ BP'' statement; it is a careful limit theorem hedged by three structural conditions, and
we cite it only with those conditions attached. Outside that regime---nonlinear activations,
finite width comparable to depth, or settling stopped short of equilibrium---there is no guarantee
that the PC gradient equals the BP gradient, and indeed the two can differ.

\subsection{The role of the settling budget}

There is a complementary way to see when the two coincide, phrased in terms of how much settling is
allowed. If the supervision signal is infinitesimally small, or the settling is stopped almost
immediately after feedforward initialization, the activities never move substantially away from
the feedforward state, the fixed-prediction assumption holds tightly, and PC reduces to BP
\citep{song2024prospective}. Any behavior in which PC differs from BP therefore requires
\emph{substantial} settling: enough relaxation steps, and a large enough state learning rate, that
the activities genuinely reconfigure before the weights move. Conversely, a PC experiment run with
a tiny settling budget is measuring backpropagation in disguise and should be expected to show no
PC-specific effect. This is a methodological razor we wield repeatedly in \Cref{ch:experiments}:
a null result for PC is only informative if the settling was real. The open empirical question,
then---the framing for RQ1---is whether, in the regimes where PC \emph{can} differ from BP, it
actually does so to any measurable and reproducible degree on standard image-classification tasks.
A sober reading of the literature is that the advantages reported for PC have not been consistently
reproduced \citep{zahid2023curvature}, which is precisely why a fair, decounfounded comparison
(\Cref{sec:protocol,sec:exp-pcbp}) is worth doing.

\section{Generative versus Discriminative PC}\label{sec:gen-disc}

A final distinction is needed because it governs what a PC network can and cannot do, and it is
easy to assume more than the architecture delivers. The distinction is one of \emph{orientation}:
which end of the network is treated as the latent cause and which as the observation.

\subsection{Discriminative orientation}

A \emph{discriminative} PC network is oriented image$\to$label: the input layer holds the image,
the output layer holds the label, and predictions flow toward the label. This is the supervised
classifier studied for most of this thesis, and it is what the clamping conventions of
\cref{sec:so-pc} describe. Such a network classifies well, but it cannot generate: clamping a
label at the output and letting the input settle does not paint a digit, because the network was
never trained to predict pixels from a class. Doing so produces noise, not an image---a point we
verify empirically rather than assume.

\subsection{Generative orientation}

A \emph{generative} PC network reverses the orientation, label$\to$image: a low-dimensional latent
or label sits at the top and the image is predicted at the bottom, so the layer sizes run, for
example, $10 \to 256 \to 256 \to 784$. Trained with the image clamped at the bottom and the
latent free at the top, this network learns a top-down generative model of the data. The very same
settling machinery of \cref{sec:so-pc} then supports a family of capabilities for free: clamping a
class and settling the image produces per-class prototypes; clamping the visible pixels and
settling the occluded ones performs inpainting; and the residual energy of a settled input,
$\tfrac{1}{2}\sum_k\lVert\varepsilon_k\rVert^2$, serves directly as an anomaly score, since an
input the model explains poorly leaves high residual energy. That a single energy and a single
settling rule yield classification, generation, inpainting, and anomaly detection simply by
changing what is clamped is one of the conceptually attractive features of predictive coding, and
the capability demonstration in \Cref{sec:exp-gen} builds directly on this orientation distinction.
