\chapter{Related Work}\label{ch:related}

This chapter surveys the prior art against which the three contributions of this
thesis must be measured, and from that survey it establishes precisely where the
present work sits. Its purpose is positioning rather than methodology: \Cref{ch:background}
has already introduced the State-Optimization formulation of Predictive Coding (PC)
that the thesis studies, and \Cref{ch:method} will describe what was actually built.
Here the concern is narrower. A central claim of this thesis is a systems claim --- that
its fused CUDA settling kernel is, to our knowledge, the first hand-written kernel of its
kind for PC --- and a claim of priority is only as strong as the audit of prior work that
backs it. We therefore devote the bulk of this chapter to a careful, source-by-source
examination of the implementations that could plausibly contain such a kernel, of the
algorithmic family (Equilibrium Propagation) that is PC's closest relative, and of the
systems-level technique (persistent, single-launch GPU kernels) that the kernel borrows
and re-applies. We then turn to the continual-learning literature, because two of the
thesis's empirical findings --- on forgetting and on the stability--plasticity trade-off ---
can only be interpreted against the metrics and methods established there. We end with a
single section of synthesis that states plainly what is, and what is not, being claimed.

A word on tone is in order before we begin. The defining commitment of this project is
honesty about the strength of its claims, and nowhere is that more important than in a
related-work chapter that grounds a priority claim. A negative-existence claim --- ``no
prior implementation ships such a kernel'' --- can never be proved absolutely, because no
audit can index every private or unpublished codebase. We therefore frame the systems
claim narrowly, qualify it with ``to our knowledge'' throughout, and are explicit about
the prior art that the kernel's \emph{technique} draws on, claiming novelty only for its
\emph{application} to the PC settling loop. The reader should come away convinced not by
breadth of assertion but by the directness of the underlying inspection.

\section{Predictive Coding Implementations and Libraries}

Predictive Coding has, over the last half-decade, acquired a small but genuine ecosystem
of software. Before asking whether any of these tools accelerates the settling loop with a
hand-written kernel, it is worth recalling \emph{why} settling is the natural place to look
for such an effort. The defining feature of State-Optimization PC, set out in
\Cref{sec:so-pc}, is that inference is not a single feedforward pass but an iterative
relaxation: for each input the layer states $s_k$ are updated for $T$ steps along the
negative gradient of the energy $E=\tfrac12\sum_k \lVert\varepsilon_k\rVert^2$, with errors
$\varepsilon_k = s_k - \mathrm{pred}_k$ and predictions $\mathrm{pred}_{i+1}=W_i\,\phi(s_i)+b_i$.
This relaxation is the computational bottleneck of PC and the part of the algorithm that
maps least gracefully onto the matrix-multiply hardware for which backpropagation was
tuned. It is, in other words, exactly where one would write a custom kernel if one were
going to. The question of this section is empirical: has anyone done so?

\subsection{The two implementation idioms: JAX/JIT and PyTorch-manual}

The mature PC libraries fall cleanly into two families distinguished by how they execute
the settling loop. The first idiom is JAX, in which the relaxation is expressed in terms of
high-level array operations and handed to the XLA compiler via \texttt{jit}, sometimes
with \texttt{vmap} for batching or an ODE solver for the continuous-time view. \citet{pinchetti2024pcx}
present PCX (the \texttt{pcax} package), a JAX framework built on \texttt{optax} that has
become a reference benchmark suite for the field; its companion paper is candid that the
JIT compiler is ``unable to parallelize the executions of the layers'' and that the
\texttt{vmap} remedy applies only ``in the unpractical case where all the layers have the
same dimension.'' \citet{millidge2024jpc} present JPC, a compact JAX/Equinox library
(under a thousand lines of core code) that casts settling as an ODE and integrates it with
Diffrax, reporting that a second-order Heun integrator settles faster than Euler. The
\texttt{ngc-learn} library of the NAC Lab and the depth-scaling work of
\citet{innocenti2026limits} (the $\mu$PC parametrisation, layered on PCX) likewise live in
JAX. In all of these, the settling loop is compiled to XLA or solved by Diffrax; there is
no separate, hand-written device kernel, and the burden of acceleration rests entirely on
the compiler.

The second idiom is PyTorch used as a tensor engine, with the settling iteration written
out explicitly as a Python loop over analytic tensor operations or, in the autograd-based
variants, as repeated calls to \texttt{backward()} on the energy. \citet{vanzwol2024preco}'s
PRECO implements PC and graph-structured PC networks this way, with the per-step state
updates expressed as manual matrix operations inside a Python \texttt{for}-$t$ loop.
Torch2PC \citep{vanzwol2024preco} computes the updates via autograd vector--Jacobian
products and offers an \texttt{Exact} mode whose gradients reproduce those of
backpropagation, which makes it a useful numerical cross-check but not an accelerated path.
The research scripts \texttt{pypc} and \texttt{pybrid} follow the same manual-tensor
pattern, as does the public code accompanying \citet{song2024prospective}'s Prospective
Configuration study. Crucially for the present thesis, the error-based reparametrisation of
\citet{goemaere2025epc} (ePC), which we discuss further in
\Cref{ch:discussion,ch:limitations} as the most closely related attack on the same end goal
of fast deep settling, is itself a PyTorch/Lightning implementation that settles by
calling \texttt{E.backward()} and stepping an optimiser. ePC accelerates PC
\emph{algorithmically} --- by optimising over the prediction errors rather than the states,
so that learning signals reach all layers at once instead of decaying exponentially with
depth --- not at the kernel level. It is therefore orthogonal to, and in principle
combinable with, a faster kernel; it must be cited and distinguished precisely so that it
is not mistaken for an overlooked instance of the systems claim.

\subsection{The audit: eleven libraries, zero settling kernels}

The grounding for this thesis's priority claim is a direct source audit of every PC library
named above, summarised here from the verification record. For each repository the audit
cloned the source (or, where rate limits intervened, inspected the file tree through the
GitHub trees API) and searched both for CUDA source files (\texttt{.cu}, \texttt{.cuh},
\texttt{.cpp}) and for the textual signatures of just-in-time kernel compilation
(\texttt{cpp\_extension}, \texttt{load\_inline}, \texttt{\_\_global\_\_}, \texttt{nvcc},
\texttt{CUDAExtension}, \texttt{triton.jit}). The result is uniform: across PCX, JPC,
\texttt{ngc-learn} (modern JAX and the TensorFlow-era \texttt{ngc-learn-legacy}), $\mu$PC,
PRECO, Torch2PC, \texttt{pypc}, \texttt{pybrid}, ePC, and Song's code, there is not a single
hand-written CUDA settling kernel. Settling runs everywhere as a Python loop over framework
operations, as a JAX/XLA JIT compilation, or as a Diffrax ODE solve --- never as one fused
custom CUDA kernel. The only apparent matches were false positives: a Python file named
\texttt{expKernel.py} in \texttt{ngc-learn} (a mathematical, not device, kernel) and the C
sources of \texttt{ngc-learn-legacy}, which turn out to be an INFIMNIST data generator
unrelated to settling. The strength of this sub-claim rests on its directness --- it is a
matter of inspection rather than inference --- and it is corroborated by PCX's own published
admission that its inference cannot be fully parallelised. This is the empirical foundation
for the first contribution's framing: the gap the kernel fills is real, and it is observable
in the source trees of the field's own tools.

\section{Equilibrium Propagation: The Closest Algorithmic Relative}

Predictive Coding does not stand alone. It belongs to a broader family of energy-based
learning algorithms whose common shape is a two-phase procedure: relax a network of units
to a fixed point of some energy function, then update the weights using only quantities
local to that fixed point. The closest relative of PC within this family is Equilibrium
Propagation (EP), introduced by \citet{scellier2017ep}. EP, like PC, settles a network to
equilibrium (its ``free phase''), then nudges the outputs toward the target and settles
again (its ``nudged phase''), and the difference between the two equilibria furnishes a
local, backprop-approximating weight update. The structural parallel to PC's
inference-then-plasticity schedule is exact enough that any audit underwriting a
``first for PC'' kernel claim must also rule out the EP side: if some EP implementation
already shipped a hand-written settling kernel, the claim would be substantially weakened
by analogy, even though EP is a different algorithm. This section reports the second half
of the audit, which closes that gap.

\subsection{The canonical and recent EP implementations}

The audit enumerated the canonical EP lineage and its more recent descendants. The original
implementation of \citet{scellier2017ep} is written in Theano. The influential variants ---
the symmetric-nudging and holomorphic refinements associated with \citet{laborieux2021ep},
the recurrent-backprop-equivalence work of Ernoult and colleagues, and the agnostic-EP and
spiking-EP reimplementations --- are written in PyTorch with the settling phases driven by
autograd, by Euler iteration of explicit tensor operations, or by a Python host loop.
\citet{laborieux2021ep}'s holomorphic variant moves to JAX with Haiku but still settles in
an eager \texttt{while} loop rather than a compiled scan or a Pallas/Triton kernel. The
recent end of the literature (for example the 2025--2026 stochastic-EP and
intermediate-error-signal scaling papers) is explicit in its own text about running ``in
Python using the PyTorch framework'' on commodity GPUs. Across roughly two dozen repositories
spanning the canonical authors, community reimplementations, and adjacent energy-based
models (modern Hopfield networks, the Energy Transformer), the kernel signatures
(\texttt{\_\_global\_\_}, \texttt{cpp\_extension}, \texttt{load\_inline}, \texttt{triton.jit})
return empty or false-positive (an EDA-placement tool and a lattice-QCD project that happen
to use the string \texttt{eqprop}).

\subsection{The single near-miss, and why it does not refute the claim}

One repository deserves to be singled out because it comes closest. A multi-algorithm
bio-plausible-learning framework (\texttt{autonull/bioplausible}) does ship hand-written GPU
kernels for EP dynamics. It does not, however, satisfy any of the three features the thesis
claims. First, it is EP, not PC --- a related but distinct algorithm, so ``first for PC''
remains untouched. Second, its kernels are written in Triton, not hand-written CUDA; the
repository contains no \texttt{.cu}/\texttt{.cuh} sources and no \texttt{\_\_global\_\_} or
\texttt{cpp\_extension} usage. Third, and most importantly, the kernels are not a fused,
persistent, states-resident in-kernel settling loop: each kernel computes a single
elementwise relaxation step of the form $h \leftarrow (1-a)\,h + a\,\tanh(\text{pre-activation})$,
the matrix multiply is \emph{not} fused in, and the settling iteration itself runs on the
host in a Python loop. (The audit also flagged a naming trap in the same repository: a
\texttt{fused\_settle\_step} living under a \texttt{cuda} path whose body is in fact pure
PyTorch operations, with a docstring that concedes ``PyTorch still launches separate
kernels.'') This repository is the nearest EP-side prior art and should be cited
defensively, but it weakens nothing: it is the wrong algorithm, the wrong implementation
technology, and the wrong execution structure. The conclusion of the EP audit is therefore
that the closest algorithmic relative \emph{hardens} rather than refutes the priority claim:
to our knowledge no EP implementation ships a hand-written CUDA settling kernel either.

\subsection{A scope caveat: analog and neuromorphic acceleration}

For completeness, and in the interest of not overstating the gap, it must be said that EP
\emph{does} have a vigorous acceleration literature --- it simply lives in physical, not
digital, hardware. Analog-circuit, coupled-oscillator, memristor, and spiking
implementations exploit EP's defining property that the free-phase relaxation is itself a
physical energy-minimisation process, with reported speed-ups of two orders of magnitude.
This is precisely the regime in which energy-based learning is most naturally at home, and
it is worth acknowledging plainly. It is not, however, prior art for a GPU-kernel claim:
these are physical devices, not software settling kernels, and the present thesis makes no
claim about them. The novelty being asserted is specific to running the PC settling loop on
a conventional digital GPU.

\section{Systems for Iterative / Fixed-Point Inference}

The argument so far establishes that no PC or EP implementation ships a hand-written
settling kernel. It would be a serious overstatement, however, to conclude that the
\emph{technique} embodied in such a kernel is novel. It is not, and intellectual honesty
requires that this be stated as forcefully as the gap itself. This section places the
kernel's design squarely within an established systems lineage and identifies what is, and
is not, the thesis's own contribution at the systems level.

\subsection{The persistent, single-launch kernel as prior art}

The pattern at the heart of an efficient settling kernel --- keep the network's state
resident on-chip and run the entire iterative loop inside a single kernel launch, rather
than issuing one launch per step --- is canonical GPU-engineering prior art, and it predates
this work by nearly a decade. Its archetype is the persistent recurrent-neural-network
kernel of \citet{diamos2016persistent}, also exposed in cuDNN as the
\texttt{CUDNN\_RNN\_ALGO\_PERSIST\_STATIC}/\texttt{DYNAMIC} algorithms. There a single launch
keeps its thread blocks resident across all recurrent timesteps, holds the weights and
activations in the streaming-multiprocessor register file and shared memory, and
synchronises between steps with a global barrier --- a design motivated explicitly by the
observation that per-step kernel-launch overhead dominates at small batch (the persistent
kernel reports order-of-magnitude speed-ups over a per-step cuBLAS baseline at mini-batch
four). The same engineering pattern has reappeared more recently in the LLM ``megakernel''
literature, in which an entire forward or decode pass is fused into one persistent,
state-resident kernel to win back small-batch launch overhead. The thesis's kernel is the
same pattern; it borrows it deliberately. What it claims is the \emph{first transfer of that
pattern to the PC settling loop} --- together with the specific layer-parallel,
$T$-sequential decomposition and the precision-weighted, $\phi'$-gated bidirectional update
that the PC dynamics require --- not the invention of the pattern.

\subsection{Deep Equilibrium models and the conceptual neighbours}

A second relevant systems lineage is that of fixed-point models that, like PC, define their
output as the solution of an equilibrium condition rather than the result of a fixed
feedforward depth. Deep Equilibrium models \citep{bai2019deq} are the canonical example:
they solve for a fixed point of a layer using root-finders such as Anderson acceleration or
Broyden's method, and back-propagate through the equilibrium using the implicit function
theorem. DEQ is the nearest \emph{conceptual} relative on the fixed-point axis, and ePC
\citep{goemaere2025epc} is the nearest relative on the energy-minimisation axis; both attack
the same end goal of fast fixed-point inference. Neither, however, does so with a
hand-written CUDA settling kernel --- DEQ solves entirely in PyTorch, and ePC, as noted
above, reparametrises the problem rather than rewriting the kernel --- so neither refutes the
systems claim. They do, however, define the conceptual neighbourhood in which the
contribution should be read.

\subsection{The framework tax: a generic observation, not a contribution}

Finally, the empirical phenomenon that the kernel exploits --- that an inference loop built
from many small framework operations becomes launch-overhead-bound at small and medium batch
sizes --- is itself a generic, already-published observation about deep-learning frameworks,
sometimes called the ``framework tax.'' PCX's own paper alludes to it when it laments the
cost of ``multiple smaller operations.'' This thesis measures that effect for the specific
case of PC settling and uses it to motivate the kernel, but it does not and must not claim
the observation itself as a contribution. The honest framing, which \Cref{ch:discussion}
develops, is that the kernel is a faithful \emph{measurement} and \emph{first application} of
a known technique to a known kind of bottleneck, not the discovery of either. This careful
attribution is what allows the priority claim to survive scrutiny: it is narrow,
PC-specific, qualified, and explicit about its debts.

\section{Continual Learning}

The thesis's empirical contributions are not confined to the kernel. Following the fair
PC-versus-backpropagation methodology that \Cref{sec:protocol} formalises, two of its
findings concern \emph{continual learning} --- the setting in which a model must learn a
sequence of tasks without revisiting earlier data. This setting is where Predictive Coding
is most often conjectured to hold an advantage over backpropagation, and so it requires its
own grounding in prior work, both to define the metrics by which any such advantage would be
measured and to position the baselines against which it is compared.

\subsection{Catastrophic forgetting and the stability--plasticity trade-off}

The core difficulty of continual learning is \emph{catastrophic forgetting}: when a network
trained by gradient descent learns a new task, the same weight updates that fit the new data
overwrite the representations supporting earlier tasks. This is one face of a more general
tension known as the stability--plasticity trade-off. A model must be \emph{plastic} enough
to acquire new tasks and \emph{stable} enough to retain old ones, and methods that improve
retention frequently do so by reducing the rate at which new tasks are learned. This
trade-off is not a confound to be eliminated but a curve along which methods sit, and it is
essential to interpreting any claim that one learning rule ``forgets less'' than another:
forgetting less while also learning each task less well is movement \emph{along} the curve,
not a strict improvement \emph{over} it. The conjectured appeal of PC in this setting,
articulated most forcefully by \citet{song2024prospective}, is that its inference phase
settles the hidden units into a ``prospective'' configuration before plasticity acts, which
is argued to distribute the error in a way that avoids the weight interference responsible
for forgetting --- the difference between altering activity to anticipate a side effect and
back-propagating a shared-weight update that damages a correct, unrelated association.

\subsection{Measuring forgetting: the GEM metrics}

To compare methods on this axis at all, one needs agreed metrics, and the standard set comes
from the Gradient Episodic Memory work of \citet{lopezpaz2017gem}. From the matrix of
test accuracies $R_{ij}$ --- the accuracy on task $j$ after training through task $i$ --- one
derives \emph{average accuracy} (overall performance at the end of training),
\emph{backward transfer} (BWT, the influence of later learning on earlier tasks, whose
negative is the amount forgotten), and \emph{forward transfer} (the influence of earlier
learning on yet-unseen tasks). These metrics are what make a statement such as ``PC forgets
less'' precise and falsifiable, and the thesis's continual-learning experiments report them
directly. \citet{lopezpaz2017gem}'s GEM itself is a replay-style method that stores a small
episodic memory of past examples and constrains new gradients not to increase the loss on
that memory; it is included here as the origin of the evaluation framework rather than as a
baseline the thesis re-implements. The broader infrastructure of the field --- standardised
benchmark streams such as Split- and Permuted-MNIST and the metric implementations that
accompany them --- is consolidated in the Avalanche library \citep{lomonaco2021avalanche},
which the thesis draws on for benchmark construction even where it computes the metrics
independently.

\subsection{Regularisation baselines: Elastic Weight Consolidation}

The most directly relevant baseline is a regularisation method rather than a replay method.
Elastic Weight Consolidation (EWC) of \citet{kirkpatrick2017ewc} reduces forgetting without
storing past data by adding a quadratic penalty that anchors each weight to its
post-task value, with a per-weight stiffness given by the diagonal of the Fisher information
matrix --- in effect, weights that mattered for previous tasks are made expensive to move.
EWC is the canonical answer to ``how do we forget less while staying within standard
backpropagation,'' and it therefore makes the natural reference point for any claim that
PC's inference phase confers a forgetting advantage. Reading PC's behaviour against EWC ---
and against a fair, loss-matched backpropagation baseline --- is what keeps the
stability--plasticity trade-off honest: it situates PC on the same curve as an explicit
anti-forgetting mechanism rather than against a naive, untuned baseline. \Cref{ch:experiments}
reports the resulting comparison, and \Cref{ch:discussion} interprets it in light of exactly
the trade-off articulated here.

\section{Positioning of This Work}

Against this backdrop the position of the thesis can be stated in a single, deliberately
modest paragraph. This is a from-scratch, reproducible study of State-Optimization PC on
MNIST-scale data, contributing three focused and independently verified results rather than
a new algorithm or a frontier-scale system. First, a systems contribution: a hand-written,
fused, single-launch CUDA settling kernel that holds the layer states resident across all
$T$ steps and that is, to our knowledge, the first such kernel for the PC settling loop ---
a claim grounded in the source audit of \Cref{ch:related} of eleven PC libraries and the
Equilibrium Propagation family, and bounded explicitly by the persistent-RNN and megakernel
prior art whose technique it re-applies rather than invents. Second, a methodological and
empirical contribution: a fair, loss-matched, multi-seed comparison of PC against
backpropagation, whose headline finding is largely \emph{null} --- vanilla PC neither clearly
beats nor clearly trails a fairly tuned, loss-matched backpropagation baseline at this scale
--- and whose value lies in the methodology and in disentangling three confounds (the loss
function, the stability--plasticity trade-off, and an untuned baseline) that can manufacture
spurious wins in either direction. Third, a set of demonstrations of PC's distinctive
capabilities --- generative reconstruction, inpainting, and anomaly detection --- and of its
behaviour under variants such as incremental PC and continual-learning regularisers. What is
\emph{not} claimed is as important as what is. The thesis does not claim to solve the
field's open problems --- the depth-scaling failure, the PC-equals-backprop puzzle, or the
scale gap \citep{innocenti2026limits} --- which remain frontier-laboratory work that a
single-author MNIST study can reproduce and characterise but not resolve. It does not claim a
general-purpose kernel, a general speed-up, novelty for the persistent-kernel technique, nor
the framework-tax observation as its own. And it does not claim a PC advantage over
backpropagation that its own honestly reported confidence intervals do not support. The
contribution is the rigour of the reproduction and the soberness of the framing; the survey
above is what licenses both.
